\PaperTitle

### 2020-05-18 paper/filter_bubbles.tex
updates to the paper

Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al. (2013, Oct.),  How retailers can keep up with consumers. \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar. 12), YouTube Extremism and the Long Tail, \textit{The Atlantic}. \url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}. Retrieved on October 3, 2019.} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
\par
Understanding how recommender systems influence user behavior is important not only for characterizing the broader consequences of such systems but also for guiding their design. In this paper we develop a theoretical model of user decision-making in contexts where recommender systems are traditionally deployed. We utilize previous empirical studies that characterize how recommender systems influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that provides an explanation for these empirical results. The key insight of our model is that individual \textit{beliefs} drive the consumption choices of individuals and that recommendations serve to provide users with information that leads them to update their beliefs. However, a crucial component of our model is that individuals' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items. We use these insights to provide guidance for recommender system designers. In order to design good recommendations it becomes crucial for designers to understand user beliefs of goods they haven't consumed thus far and how these beliefs evolve over time.

@@

\xhdr{Our Contributions}
Our model provides a clear mechanism that explains the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter bubble effects even without recommendation and that recommendation leads to users being less likely to fall into such filter bubbles. In this context, filter bubble effects are defined as users consuming items in an increasingly narrow portion of the product space. The simple and intuitive driving force of this in our model is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar item. Crucially, this not only impacts the underlying belief about the average utility of the item, but also the level of uncertainty. Consequently, this learning spillover induces users to consume items ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space. This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty. However, by providing information to users, recommendation leads users to be more likely to explore other portions of the product space, limiting the filter bubble effect.
\par
We find that, while recommendation leads a single user to be more likely to explore diverse portions of the product space, it also coordinates consumption choices across users. This leads to an increase in homogeneity across users, resulting in a tradeoff between homogenizing across user consumption and diversifying within user consumption. We finally explore the relationship between the overall diversity of consumed items and user welfare. We find that more diverse sets of consumed items do not always correspond to higher user welfare. 
\par
Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems as well as the data that is traditionally collected for them. Our model highlights the importance of user beliefs in determining the consumption choices of individuals and how both recommendation and informational spillovers determine how these beliefs change over time. By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change their beliefs and resulting consumption decisions. Thus, our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without recommendation. We discuss how this provides an additional rationale as to why ``accurate" recommendations are not always good recommendations.
\par

@@

Another set of papers has examined whether recommender systems can lead users to become increasingly homogenized or consume similar content to one another. \cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization. \cite{chaney2018algorithmic} show how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations. \cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories. Our model leads to similar results as previous theoretical and empirical work where recommender systems lead to increased user homogenization. However, the mechanisms behind this from our model differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.

@@

Another set of papers in the recommender systems community that studies the impact of human decision-making on the design and evaluation of recommender systems. \cite{chen2013human} surveys the literature on the relationship between human decision making and recommender systems. The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process. Our model points out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of them and both consumption and recommendation provide them with information to reduce their uncertainty. Thus, the primary insight of our paper is that individual beliefs and how individuals update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of recommender systems. Within this literature, there is a set of papers that focus on ``user-centric" approaches to recommendation \cite{celma2008new, cremonesi2013user, pu2011user} whereby user evaluation of the usefulness of recommendation is a key evaluation measure. Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs. Finally, \cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.

>>>>

Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al. (2013, Oct.),  How retailers can keep up with consumers. \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
\par
Understanding how recommender systems influence user behavior is important not only for characterizing the broader consequences of such systems but also for guiding their design. In this paper we develop a theoretical model of user decision-making in contexts where recommender systems are traditionally deployed. We utilize previous empirical studies that characterize how recommender systems influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that provides an explanation for these empirical results. The key insight of our model is that individual \textit{beliefs} drive the consumption choices of individuals and that recommendations serve to provide users with information that leads them to update their beliefs. However, a crucial component of our model is that individuals' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items. We use these insights to provide guidance for recommender system designers. In order to design good recommendations it becomes crucial for designers to understand user beliefs of goods they haven't consumed thus far and how these beliefs evolve over time.

@@

\xhdr{Our Contributions}
Our model provides a clear mechanism that explains the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter bubble effects even without recommendation and that recommendation leads to users being less likely to fall into such filter bubbles. In this context, filter bubble effects are defined as users consuming items in an increasingly narrow portion of the product space. The simple and intuitive driving force of this in our model is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar items. Crucially, this not only impacts the underlying belief about the average utility of the item, but also the level of uncertainty. Consequently, this learning spillover induces users to consume items ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space. This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty. However, by providing information to users, recommendation leads users to be more likely to explore other portions of the product space, limiting the filter bubble effect.
\par
We find that, while recommendation leads a single user to be more likely to explore diverse portions of the product space, it also coordinates consumption choices across users. This leads to an increase in homogeneity across users, resulting in a tradeoff between homogenizing across user consumption and diversifying within user consumption. We finally explore the relationship between the overall diversity of consumed items and user welfare. We find that more diverse sets of consumed items do not always correspond to higher user welfare. 
\par
Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems as well as the data that is traditionally collected for them. Our model highlights the importance of user beliefs in determining the consumption choices of individuals and how both recommendation and informational spillovers determine how these beliefs change over time. By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change their beliefs and resulting consumption decisions. Thus,  our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without it. We discuss how this provides an additional rationale as to why ``accurate" recommendations are not always good recommendations.
\par

@@

Another set of papers has examined whether recommender systems can lead users to become increasingly homogenized or consume similar content to one another. \cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization. \cite{chaney2018algorithmic} shows how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations. \cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories. Our model leads to similar results as previous theoretical and empirical work where recommender systems lead to increased user homogenization. However, the mechanisms behind this from our model differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.

@@

Another set of papers studies the impact of human decision-making on the design and evaluation of recommender systems. \cite{chen2013human} surveys the literature on the relationship between human decision making and recommender systems. The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process. Our model points out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of them and both consumption and recommendation provide them with information to reduce their uncertainty. Thus, the primary insight of our paper is that individual beliefs and how individuals update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of recommender systems. Within this literature, there is a set of papers that focus on ``user-centric" approaches to recommendation \cite{celma2008new, cremonesi2013user, pu2011user} whereby user evaluation of the usefulness of recommendation is a key evaluation measure. Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs. Finally, \cite{hodgson2019horse} considers a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
